---
title: "Appendix 1. Record linkage from the certification roster to the national provider registry"
subtitle: "Supplemental Digital Content"
format:
  docx:
    # reference-doc: templates/green_journal_template.docx
    number-sections: false
  html:
    toc: true
    embed-resources: true
bibliography: references.bib
csl: green-journal.csl
params:
  dev_mode: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- if (dir.exists("artifacts")) "." else ".."
source(file.path(root, "manuscript", "R", "build_stats_catalog.R"))
source(file.path(root, "manuscript", "R", "inline_stats.R"))
if (isTRUE(params$dev_mode)) Sys.setenv(MANUSCRIPT_DEV_MODE = "1") else
  Sys.unsetenv("MANUSCRIPT_DEV_MODE")

# WIRED TO THE SAME CATALOG AS THE MAIN DOCUMENT, deliberately.
# ~/isochrones/manuscript/supplement_GREEN_JOURNAL_WORD.Rmd renders independently
# and re-declares its constants, with a long comment explaining that they MIRROR
# the main document's catalog values. Mirroring is the thing that drifts: it is
# the same defect as a number typed into a paragraph, one indirection further
# out. The supplement and the manuscript here read one catalog, so a supplement
# figure cannot disagree with the body that cites it.
K <- mw_init_stats(root)
```

# eMethods 1. Why linkage is the binding constraint

The certification roster publishes credential, status and dates for every
certificant and **no address at any geographic level**. It carries no National
Provider Identifier, no date of birth, and no practice affiliation. The
provider registry carries practice address for every enumerated clinician but
no certification number. The two must therefore be joined on name, and every
geographic statement in the main analysis rests on that join.

Name is a weak key in this population. Certified nurse-midwives are
overwhelmingly female (`r mw_safe_stat("table1.female_pct", "%.1f")`% of the
linked cohort), a population with a high rate of surname change, and the
registry's historical files record whatever name was current in each annual
snapshot. Middle name is absent from the roster entirely and is missing from a
large share of registry records.

# eMethods 2. Ranked evidence classes, and the fail-closed rule

Candidate pairs were ranked into ordered classes of name evidence rather than
scored on a blended similarity metric. A record resolved only when **exactly
one** candidate occupied its strongest available class.

Two consequences follow, and both are intended.

**A record with several equally good candidates does not resolve.** Two
certificants with identical surname and given name, no middle name on either
side, and no date of birth are genuinely indistinguishable in these data. A
similarity score would return a winner, and the winner would be an artifact of
the scoring function rather than a fact about either person. Such records are
retained as a reported stratum, not discarded.

**Taxonomy sets the tier but never breaks a tie.** Provider taxonomy was used
to decide which class a candidate belongs to. It was never permitted to select
between two candidates within a class, because a taxonomy field describes what
a clinician is enumerated as, not which clinician a certification number
belongs to. A match supported only by nursing taxonomy is reported as its own
stratum and is never promoted into the primary cohort.

The cost of the rule is coverage:
`r mw_safe_stat("linkage.ambiguous_pct", "%.1f")`% of records
(`r mw_n("linkage.ambiguous")`) are left unresolved that a scored matcher would
have assigned. The benefit is that the resolved set contains no manufactured
identities.

# eResults 1. Disposition of every record

```{r tblS1}
d <- K$linkage$table
disp <- setdiff(names(d), c("status", "n", "pct_matched"))
lab <- c(matched = "Resolved, midwifery taxonomy",
         matched_nursing_taxonomy = "Resolved, nursing taxonomy only",
         ambiguous_tied_names = "Unresolved, tied names",
         ambiguous_contested_npi = "Unresolved, contested identifier",
         ambiguous_unruled_out_component = "Unresolved, component not ruled out",
         unmatched = "Unresolved, no plausible candidate",
         candidate_class5_held_out_of_cohort = "Held out of cohort by design")
tot <- sum(d$n)
t <- data.frame(Disposition = ifelse(is.na(lab[disp]), disp, lab[disp]),
                No. = vapply(disp, function(cn) sum(d[[cn]]), numeric(1)),
                check.names = FALSE)
t$`%` <- sprintf("%.1f", 100 * t$No. / tot)
t <- t[order(-t$No.), ]
t$No. <- formatC(t$No., format = "d", big.mark = ",")
t <- rbind(t, data.frame(Disposition = "Total",
                         No. = formatC(tot, format = "d", big.mark = ","),
                         `%` = "100.0", check.names = FALSE))
knitr::kable(t, row.names = FALSE, align = "lrr",
  caption = sprintf("Table S1. Disposition of all %s certification records. The dispositions are exhaustive and mutually exclusive; they sum to the total by construction rather than by arithmetic performed after the fact.",
                    formatC(tot, format = "d", big.mark = ",")))
```

Of `r mw_n("linkage.total")` records, `r mw_n("linkage.matched")`
(`r mw_safe_stat("linkage.matched_pct", "%.1f")`%) resolved to a single
provider with confirmed midwifery taxonomy and `r mw_n("linkage.nursing")`
(`r mw_safe_stat("linkage.nursing_pct", "%.1f")`%) on nursing taxonomy alone,
giving `r mw_safe_stat("linkage.any_tier_pct", "%.1f")`% resolved on any
accepted tier. `r mw_n("linkage.ambiguous")`
(`r mw_safe_stat("linkage.ambiguous_pct", "%.1f")`%) had plausible candidates
that could not be distinguished, `r mw_n("linkage.unmatched")`
(`r mw_safe_stat("linkage.unmatched_pct", "%.1f")`%) had no plausible
candidate at all, and `r mw_n("linkage.heldout")`
(`r mw_safe_stat("linkage.heldout_pct", "%.1f")`%) were held out by design.

The distinction between the last two is worth preserving. *No plausible
candidate* and *several indistinguishable candidates* are different states of
knowledge: the first is consistent with a clinician who never enumerated, the
second with one who did and whose record cannot be told from a colleague's.
Collapsing them into a single "unmatched" category, which is the common
practice, would discard the information that separates a coverage problem from
an identity problem.

# eResults 2. Linkage is selected on certification status

```{r tblS2}
d <- K$linkage$table
t2 <- data.frame(
  `Certification status` = d$status,
  Records = formatC(d$n, format = "d", big.mark = ","),
  Resolved = formatC(d$matched, format = "d", big.mark = ","),
  `Resolved, %` = sprintf("%.1f", d$pct_matched),
  check.names = FALSE)
knitr::kable(t2, row.names = FALSE, align = "lrrr",
  caption = "Table S2. Resolution on midwifery taxonomy by certification status, ordered by number of records.")
```

Resolution ranges from `r mw_safe_stat("linkage.active_pct", "%.1f")`% among
ACTIVE certificants to `r mw_safe_stat("linkage.dead_pct", "%.1f")`% among
DECEASED certificants. The gradient is mechanical rather than mysterious: the
registry is a registry of clinicians who enumerated to bill, enumeration became
near-universal only over the study period, and a clinician who left practice
before enumeration was routine has no registry record to be linked to.

**This is the central threat to validity in the main analysis, and its
direction is knowable.** The persistence estimates describe providers who both
linked and remained observable in the annual snapshots. Providers who left the
workforce are systematically less likely to be in that set. Because leaving the
workforce and leaving a county are positively associated -- a clinician who
retires often stops appearing in the registry entirely rather than appearing at
a new address -- the observed cohort is enriched for people who stayed, in both
senses. **The reported persistence figures are therefore upper bounds on
persistence among all certificants.**

For the PERSISTENCE estimates the magnitude of that bias cannot be quantified
from these data. Persistence requires repeated observation, and the
certificants outside the cohort have none by construction; no arithmetic
recovers a trajectory that was never observed. Narrowing that bound would
require linking the roster to a source that does not require enumeration -- a
state licensure file, or a death index for the DECEASED stratum -- which is the
natural next step and is not attempted here.

For the COMPOSITION estimates it can. An earlier version of this appendix said
otherwise, and that was too strong: rurality is missing because a practice ZIP
code failed to resolve, not because a certificant failed to enter the cohort,
and those are different events. Placing every certificant with no assignable
county entirely inside and then entirely outside each rurality category bounds
the roster-wide metropolitan share between
`r mw_safe_stat("bounds.lower_pct", "%.1f")`% and
`r mw_safe_stat("bounds.upper_pct", "%.1f")`% with no assumption about the
missingness mechanism at all. Discarding the
`r mw_n("bounds.outside_n")` non-cohort certificants whose ZIP code does
resolve would widen that to
`r mw_safe_stat("bounds.manski_lower_pct", "%.1f")`%--`r mw_safe_stat("bounds.manski_upper_pct", "%.1f")`%,
which is a wider interval bought by ignoring evidence rather than by being more
careful. Restricting to ACTIVE certificants, the population a workforce claim
usually means, gives
`r mw_safe_stat("bounds.active_lower_pct", "%.1f")`%--`r mw_safe_stat("bounds.active_upper_pct", "%.1f")`%.
Under the further assumption that certification status explains the
missingness, inverse-probability weighting returns
`r mw_safe_stat("bounds.ipw_pct", "%.1f")`%; it is reported as a sensitivity
and not as a correction, because status plainly does not explain all of it.

# eResults 3. What the unresolved records are not

`r mw_n("linkage.ambiguous")` unresolved records are not a random sample of the
roster. Tied names concentrate in common surnames, which are unevenly
distributed across ancestry groups, and any analysis stratifying on race,
ethnicity or national origin using this cohort would inherit that
non-randomness. The main analysis does not stratify on those attributes, which
is one reason it does not report them.

# eMethods 3. Reproducibility

The linkage is frozen: a fixed crosswalk file with a recorded SHA-256
checksum, not a matcher re-run at analysis time. Every artifact in this
appendix is written with a provenance record listing the checksum of each of
its inputs, and the numbers in this document are recomputed from those
artifacts at render time rather than transcribed. Code and artifacts are
archived[@midwifery_repo]; the person-level crosswalk itself is not
redistributed.

# References

::: {#refs}
:::
