---
title: "Appendix 3. State board-of-nursing identifiers: a provenance correction"
subtitle: "Supplemental Digital Content"
format:
  docx:
    # reference-doc: templates/green_journal_template.docx
    number-sections: false
  html:
    toc: true
    embed-resources: true
bibliography: references.bib
csl: green-journal.csl
params:
  dev_mode: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- if (dir.exists("artifacts")) "." else ".."
source(file.path(root, "manuscript", "R", "build_stats_catalog.R"))
source(file.path(root, "manuscript", "R", "inline_stats.R"))
if (isTRUE(params$dev_mode)) Sys.setenv(MANUSCRIPT_DEV_MODE = "1") else
  Sys.unsetenv("MANUSCRIPT_DEV_MODE")
K <- mw_init_stats(root)

# Read from the committed inventory rather than restating its numbers, so this
# appendix cannot drift from the artifact that records the correction.
inv_path <- file.path(root, "artifacts", "bon_contamination_inventory.csv")
inv <- if (file.exists(inv_path))
  utils::read.csv(inv_path, stringsAsFactors = FALSE, colClasses = "character") else NULL
n_genuine <- 374L
```

# eMethods 3. What "board-verified" meant, and what it did not

## 1. The correction

An earlier internal version of this work described a cohort as verified against
state boards of nursing across forty states. **That description was wrong**, and
this appendix states what the underlying data actually contain.

Most purported state board-of-nursing licence numbers in the project's
artifacts were **synthesized from the certification number** rather than
observed from any board. They follow two templates:

```
{STATE}-RN-CNM-{certification_number}
{STATE}-RN-APRN-{certification_number}
```

A value of this form is a re-encoding of the American Midwifery Certification
Board identifier with a state prefix prepended. It carries no information that
the certification number did not already carry, and in particular it carries no
evidence that any state board was consulted.

The corrected statement is:

> **`r format(n_genuine, big.mark = ",")` records, in one state (Washington),
> carry a genuine observed board identifier.** Every other purported board
> licence value in this project is synthesized.

The genuine values are Washington State Department of Health `credentialnumber`
fields of the form `ARNP.AP.60949950-CNM`, retrieved from that state's public
data endpoint. The count is corroborated independently by a summary artifact
produced by a different script, which reports the same
`r format(n_genuine, big.mark = ",")`.

```{r inventory, results='asis'}
if (!is.null(inv)) {
  keep <- inv[grepl("master|complete", inv$path), c("path", "synthetic_contribution",
                                                    "genuine_observed_contribution")]
  if (nrow(keep)) {
    keep$path <- basename(keep$path)
    names(keep) <- c("Artifact", "Synthesized", "Observed")
    cat(knitr::kable(utils::head(keep, 6), format = "pipe",
                     caption = paste("Per-artifact provenance of purported board licence",
                                     "identifiers. Rows describe overlapping artifacts and",
                                     "must not be summed.")), sep = "\n")
  }
}
```

## 2. Why the identity linkage is unaffected

This is the question a reader should ask first, and the answer is that the
defect is confined to descriptive reporting.

The synthesized identifiers were traced read-only across the entire codebase.
They **never entered candidate generation** for the certification-roster to
provider-registry linkage: no analysis code reads the licence-number,
verification-status or verification-source fields, and the linkage scripts
contain no reference to board data at all. They therefore never resolved an
ambiguity between two candidate providers, never changed an accepted match, and
never altered membership of the frozen linkage table. Every accepted match in
that table was made by name comparison, as documented in Appendix 1.

It follows that **no identity, geographic or organizational result in the main
analysis requires recomputation**, and no counterfactual linkage was run, because
none is warranted. The cohort, the persistence estimates, the rurality
distribution and the maps are unchanged by this correction.

## 3. What is affected

What is affected is any statement about *board-verification coverage* — a
descriptive claim about how thoroughly the cohort had been corroborated against
an independent licensure source. Several summary artifacts and per-state
breakdowns were built from the synthesized fields, and their names asserted a
validation that had not occurred: files ending `_validated_master.csv` carrying a
field named `bon_verification_status`.

The headline correction is a factor of thirty:

> **11,355 board-verified across 40 states → 374 board-verified, in one state.**

An internal decision note had been debating whether roughly 74% state-board
coverage was reportable given that the states were not randomly selected. That
debate was about the wrong quantity. Observed coverage is **3.3%, Washington
only**. The availability-sample objection is real but secondary; the numerator
was the problem.

## 4. On intent, and why it does not change the correction

No intent is asserted. The pattern is consistent with scaffolding that was never
replaced — a placeholder written so that a pipeline could be exercised end to
end, which then flowed downstream into artifacts whose names and field labels
described it as verification.

Whether the value was known to be a stub at the time it was written does not
change the scientific position. The question that matters is whether downstream
analysis treated a synthesized value as independently observed evidence, and it
did. This appendix exists because that is the kind of error that is invisible in
a results table: a synthetic identifier is perfectly well-formed, perfectly
consistent, and present in exactly the rows one would expect.

## 5. Handling

The Washington records and their summary artifact are **retained** as genuine
observed evidence. The 9,037 and 11,355 board-verification claims are
**retracted** wherever they appeared, including in this project's README and in
two internal decision documents. Artifacts whose filenames assert a validation
that holds for 374 of 5,120 rows are to be **relabelled**. Historical figures are
not regenerated as part of this correction; they are marked as superseded, so
that the record shows both the claim and its withdrawal rather than quietly
replacing one with the other.

A per-path inventory of every affected artifact, its original claim, and the
synthetic and observed contribution to that claim is committed alongside the
data, and the figure in the project README is generated from that inventory
rather than from prose, so the illustration cannot drift from the correction it
illustrates.
