---
title: "Geographic Persistence of the Certified Nurse-Midwife and Certified Midwife Workforce in the United States, 2007--2025"
running-head: "Geographic Persistence of US Midwives"
author:
  - name: Tyler Muffly, MD
  - name: Bree Thumm, "[degrees]"
date: today
format:
  docx:
    # reference-doc: templates/green_journal_template.docx
    #
    # Uncomment for the submission render. To build the template: open a blank
    # Word document, apply the Green Journal style guide -- Times New Roman
    # 12 pt, one-inch margins, double-spaced, continuous line numbers -- and
    # save it to manuscript/templates/green_journal_template.docx. The same
    # commented-out line and the same instructions sit in
    # ~/isochrones/manuscript/manuscript_GREEN_JOURNAL_WORD_with_tmm_edits.Rmd;
    # this is the second manuscript to need it, so build it once and point both
    # files at it rather than making a third copy.
    number-sections: false
    fig-cap-location: top
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
bibliography: references.bib
csl: green-journal.csl
link-citations: true
params:
  dev_mode: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Render from the manuscript/ directory or from the repository root; resolve
# either way rather than depending on which one the caller happened to use.
root <- if (dir.exists("artifacts")) "." else ".."
source(file.path(root, "manuscript", "R", "build_stats_catalog.R"))
source(file.path(root, "manuscript", "R", "inline_stats.R"))

# DEV MODE. A missing statistic renders as a bold [PENDING] marker instead of
# stopping the build. Set params$dev_mode: false -- or render with
# MANUSCRIPT_DEV_MODE unset -- for the submission render, where a missing
# number must fail loudly rather than reach a reviewer as an empty space.
if (isTRUE(params$dev_mode)) {
  Sys.setenv(MANUSCRIPT_DEV_MODE = "1")
} else {
  Sys.unsetenv("MANUSCRIPT_DEV_MODE")
}

# REBUILT ON EVERY RENDER. Not cached, deliberately -- see the header of
# build_stats_catalog.R for the three-vintages-of-one-number problem this
# solves. If an input artifact has moved, this manuscript moves with it.
K <- mw_init_stats(root)
```

::: {.callout-note appearance="simple"}
Statistics catalog rebuilt `r K$meta$built` under R `r K$meta$r_version`.
Every number below is computed at render time from the artifacts named in
`manuscript/R/build_stats_catalog.R`. None is typed into this document.
:::

# Abstract

**OBJECTIVE:** To estimate how often certified nurse-midwives and certified
midwives remain in the same county over annual and career horizons, and whether
persistence differs by the rurality of the county in which they were first
observed practicing.

**METHODS:** Retrospective cohort study using only publicly available records.
The American Midwifery Certification Board verification directory
(`r mw_n("linkage.total")` certification records) was linked to the National
Plan and Provider Enumeration System using ranked name-evidence classes,
resolving a record only when exactly one candidate occupied its strongest
available class. The cohort was followed across
`r mw_safe_stat("panel.snapshots")` consecutive annual provider snapshots
(`r mw_safe_stat("panel.year_min")`--`r mw_safe_stat("panel.year_max")`).
County was assigned from practice ZIP through a ZCTA-to-county crosswalk and,
in sensitivity analysis, through a geocoding cascade with point-in-polygon
assignment. Annual persistence was the same county in consecutive snapshots;
career persistence the same county at first and last observation. Origin
rurality was the Rural-Urban Continuum Code of the first observed county.
Proportions carry Wilson score intervals; ordered strata were compared with a
Cochran-Armitage test for trend.

**RESULTS:** `r mw_n("linkage.matched")`
(`r mw_safe_stat("linkage.matched_pct", "%.1f")`%) of records resolved to a
single provider with midwifery taxonomy; `r mw_n("linkage.ambiguous")`
(`r mw_safe_stat("linkage.ambiguous_pct", "%.1f")`%) were left unresolved as
indistinguishable. The cohort comprised `r mw_n("panel.cohort_n")` providers
contributing `r mw_n("panel.provider_years")` provider-years (median
`r mw_safe_stat("panel.median_years")` years observed). Annual same-county
persistence was
`r mw_pct_ci(mw_stat("persist.annual_county"), mw_stat("persist.annual_county_lo"), mw_stat("persist.annual_county_hi"))`
across `r mw_n("panel.pairs_used")` consecutive-year pairs. Career same-county
persistence, over a median `r mw_safe_stat("panel.median_span")`-year span, was
`r mw_safe_stat("persist.career_county", "%.1f")`%. By origin rurality it was
`r mw_pct_ci(K$rural$table$pct[1], K$rural$table$lo[1], K$rural$table$hi[1])`
for metropolitan,
`r mw_pct_ci(K$rural$table$pct[2], K$rural$table$lo[2], K$rural$table$hi[2])`
for nonmetropolitan adjacent, and
`r mw_pct_ci(K$rural$table$pct[3], K$rural$table$lo[3], K$rural$table$hi[3])`
for nonmetropolitan remote counties (`r mw_pval(K$rural$trend_p)` for trend;
metropolitan minus remote `r mw_safe_stat("rural.diff", "%.1f")` percentage
points, 95% CI `r mw_safe_stat("rural.diff_lo", "%.1f")`--`r mw_safe_stat("rural.diff_hi", "%.1f")`).
Rural-origin midwives who moved went predominantly to metropolitan counties.

**CONCLUSION:** Retention of certified midwives is geographically sticky over
one to three years but materially less so across a career. Models projecting
county-level access gains from retention interventions should carry an explicit
mobility term at decade horizons.

# Introduction

Whether a pregnant person can reach a midwife depends not on how many midwives
have been certified but on where they practice relative to where births occur.
Access to midwifery care is a distributed quantity. Rural counties have lost
obstetric services steadily over the past two decades[@ref42], and rural
hospital closures have accelerated that loss[@Kaufman2016rural], so the
distribution of the remaining perinatal workforce has direct consequences for
who can reach care[@ref6]. Yet the national midwifery workforce literature has
necessarily characterized midwives as a population rather than as a
distribution[@todo_midwife_supply], because their locations have not been
available to characterize.

The reason is structural. The authoritative roster of certified midwives in the
United States publishes certification number, credential, status, and dates for
every certificant. It publishes no address at any geographic level: not county,
not ZIP code, not state. Every geographic statement about this workforce must
therefore be derived by linking the roster to a registry that does carry
location, and the quality of that derivation governs the quality of every
downstream claim.

This constraint matters most for retention. Workforce retention is widely
proposed as a route to improving obstetric access, particularly in rural
counties losing services[@todo_midwife_attrition]. That argument contains an
unmeasured geographic assumption: that a midwife retained in the profession is
a midwife retained *in the same community*. If mobility is high, retention
preserves the national workforce while redistributing it, and a county-level
access gain cannot be assumed from a national retention effect. If mobility is
low, retention is geographically sticky and the incidence of a retention effect
falls approximately where the affected midwives already practice. We are aware
of no national estimate of this quantity for certified midwives. Our objective
was to estimate it directly, at both short and career horizons, and to test
whether it differs by the rurality of the county where a midwife began
practice.

# Methods

## Data sources

All sources were publicly available; no proprietary vendor file, purchased
roster, or restricted-access dataset was used. The certification roster came
from the American Midwifery Certification Board public verification
directory[@amcb2026]. Provider identity and practice address came from the
National Plan and Provider Enumeration System, using the historical
dissemination files for each year from `r mw_safe_stat("panel.year_min")`
through `r mw_safe_stat("panel.year_max")`[@nppes], a complete series with no
missing year. County boundaries were the 2023 Census cartographic boundary
files[@census_cb2023]; ZIP-to-county correspondence was the Census 2020
ZCTA-to-county relationship file[@census_zcta]; rurality was the 2023 US
Department of Agriculture Rural-Urban Continuum Code[@rucc2023].

The study population was midwives certified by the American Midwifery
Certification Board, comprising certified nurse-midwives and certified
midwives. Certified professional midwives and state-licensed direct-entry
midwives are certified by other bodies, are not enumerated in this directory,
and are outside the prespecified study population; they are not included in any
denominator, analysis, or interpretation.

## Record linkage

Because neither source carries a shared identifier, linkage was by name over
the `r mw_safe_stat("panel.year_min")`--`r mw_safe_stat("panel.year_max")`
provider panel. Candidate pairs were ranked into ordered name-evidence classes
rather than scored on a blended similarity metric. A certification record
resolved to a provider only when exactly one candidate occupied its strongest
available evidence class. Where two or more candidates were indistinguishable
at that class -- most commonly identical surname and given name with no middle
name available on either side -- the record was left unresolved and retained as
a reported stratum rather than assigned to the most probable candidate.

This fail-closed rule trades coverage for correctness deliberately. Two
certificants with identical names, no middle name, and no date of birth are
genuinely indistinguishable in these data, and a similarity score would
manufacture certainty rather than establish it. Provider taxonomy assigned the
evidence tier but was never permitted to break an identity tie; a match
supported only by nursing taxonomy was reported as a separate stratum and never
promoted into the primary cohort.

## Cohort and follow-up

The analytic cohort was the set of providers resolved by the frozen linkage.
Each was followed across the `r mw_safe_stat("panel.snapshots")` annual
snapshots. A provider-year was one snapshot in which a provider appeared with a
usable practice address. Providers were not required to appear in every year;
entry and exit are themselves observed quantities.

## Geographic assignment

The provider panel carries practice ZIP code and state, not county. In the
primary analysis, county was assigned by mapping each ZIP to the county holding
the largest share of that ZCTA's land area, across
`r mw_n("geo.zctas")` ZCTAs. Because
`r mw_safe_stat("geo.span_multi_pct", "%.1f")`% of ZCTAs span more than one
county, this rule is consequential: `r mw_n("geo.arbitrary_zctas")` ZCTAs have a
dominant county holding under 60% of their land, and those assignments are
arbitrary. A US Postal Service ZIP is not a ZCTA, and post-office-box and
unique-business ZIPs have no ZCTA at all;
`r mw_safe_stat("geo.py_matched_pct", "%.1f")`% of provider-years matched, and
the unmatched remainder is disproportionately administrative addresses.

In sensitivity analysis, addresses were geocoded through a three-tier cascade
(Census geocoder, then ArcGIS, then centroid fallback) and assigned to county
by point-in-polygon. The cascade resolved
`r mw_safe_stat("geo.cascade_pct", "%.1f")`% of `r mw_n("geo.cascade_addresses")`
distinct addresses, and point-in-polygon placed
`r mw_safe_stat("geo.pip_pct", "%.1f")`% of the resulting
`r mw_n("geo.pip_points")` distinct points inside a county.

## Outcomes

Annual persistence was the proportion of consecutive-year pairs, among
providers observed in both years, in which county was unchanged. Career
persistence was the proportion of providers whose county at last observation
equaled their county at first observation. Both definitions were also applied
to ZIP code and to state, which bound the county estimate from below and above
and serve as an internal consistency check.

Origin rurality was the Rural-Urban Continuum Code stratum of the county of a
provider's *first* observed snapshot, banded as metropolitan (codes 1--3),
nonmetropolitan adjacent to a metropolitan area (4--6), and nonmetropolitan
remote (7--9). Classification by origin rather than by current county is
essential: classifying by current county counts a midwife who left a remote
county for a metropolitan one as metropolitan, and attributes her move to
metropolitan mobility.

## Statistical analysis

Proportions carry 95% Wilson score intervals, chosen over the normal
approximation because the remote stratum is small. Trend across the three
ordered rurality strata was tested with the Cochran-Armitage test; the
difference between extreme strata carries a normal-approximation interval.
Analyses were performed in R version `r K$meta$r_version`[@rlang]. Analysis code
and every published artifact are archived[@midwifery_repo].

## Reproducibility and ethics

This study used only publicly available records and did not involve human
subjects research; institutional review board review was therefore not
required. Every published figure is written with a provenance record listing
the SHA-256 checksum of each input file. Person-level intermediate files are
not redistributed.

Ten properties of the analysis -- among them that population is conserved, that
absent geography is never imputed, that degrading an input cannot improve the
answer, and that identical inputs produce identical outputs -- are declared in a
machine-readable registry and evaluated automatically. Each carries a
deliberately planted defect that it is required to detect, so that a check
reporting no violations has also demonstrated that it is capable of reporting
one. Appendix 2, available online at http://links.lww.com/AOG/XXXX, documents
what is verified, one defect these checks identified before it reached any
reported figure, and -- at greater length -- what such verification cannot
establish, including that it cannot detect the differential linkage described
above.

# Results

## Cohort

Of `r mw_n("linkage.total")` certification records,
`r mw_n("linkage.matched")` (`r mw_safe_stat("linkage.matched_pct", "%.1f")`%)
resolved to exactly one provider with confirmed midwifery taxonomy. Linkage was
strongly related to certification status --
`r mw_safe_stat("linkage.active_pct", "%.1f")`% of ACTIVE against
`r mw_safe_stat("linkage.dead_pct", "%.1f")`% of DECEASED certificants -- so
providers who have left the workforce are under-represented, and every estimate
below describes the *located* workforce rather than all certificants. The full
disposition of every record, the evidence classes behind it, and an analysis of
the resulting selection appear in Appendix 1, available online at
http://links.lww.com/AOG/XXXX.

The analytic cohort comprised `r mw_n("panel.cohort_n")` linked providers, of
whom `r mw_n("panel.observed")` were observed at least once. They contributed
`r mw_n("panel.provider_years")` provider-years, with a median of
`r mw_safe_stat("panel.median_years")` years observed per provider. The cohort
was `r mw_safe_stat("table1.cnm_pct", "%.0f")`% certified nurse-midwives and
`r mw_safe_stat("table1.cm_pct", "%.0f")`% certified midwives, and
`r mw_safe_stat("table1.female_pct", "%.1f")`% female. Among the
`r mw_n("cohort.known_n")` cohort members with an assignable practice county,
`r mw_n("cohort.metro_n")` (`r mw_safe_stat("cohort.metro_pct", "%.1f")`%)
practiced in metropolitan counties,
`r mw_safe_stat("cohort.adj_pct", "%.1f")`% in nonmetropolitan adjacent
counties and `r mw_safe_stat("cohort.remote_pct", "%.1f")`% in nonmetropolitan
remote counties; rurality could not be assigned for the remaining
`r mw_n("cohort.unknown_n")` (Table 1).

That distribution describes the cohort, and cohort membership differed
substantially by certification status, so it need not describe the roster of
`r mw_n("bounds.roster_n")` certificants from which the cohort was drawn.
Making no assumption about the missingness mechanism, the roster-wide
metropolitan share was bounded between
`r mw_safe_stat("bounds.lower_pct", "%.1f")`% and
`r mw_safe_stat("bounds.upper_pct", "%.1f")`%. The direction of the selection
is not, however, purely a matter of assumption: among the
`r mw_n("bounds.outside_n")` certificants outside the cohort whose practice ZIP
code nonetheless permitted a county assignment,
`r mw_safe_stat("bounds.outside_pct", "%.1f")`% were metropolitan --
`r mw_safe_stat("bounds.outside_diff_pp", "%.1f")` percentage points below the
cohort rather than above it. For the roster-wide share to fall as low as
`r mw_safe_stat("bounds.tip_threshold", "%.0f")`%, the
`r mw_n("bounds.unobserved_n")` certificants with no assignable county would
have to be only `r mw_safe_stat("bounds.tip_required", "%.1f")`%
metropolitan, `r mw_safe_stat("bounds.tip_departure", "%.1f")` percentage
points below what was observed in the cohort.

State-level supply and the midwife share of birth
attendants are mapped in Figures 3 and 4. An interactive national map of the
located cohort, showing each midwife's last observed practice location with
a deep link to the corresponding state board of nursing lookup page -- a route
to a board's own record, not an assertion that the record was checked
(Appendix 3) -- is
available at https://mufflyt.github.io/midwifery/cnm_national_leaflet_map.html[@map_cnm_national].

## Persistence

Annual same-county persistence was
`r mw_pct_ci(mw_stat("persist.annual_county"), mw_stat("persist.annual_county_lo"), mw_stat("persist.annual_county_hi"))`
across `r mw_n("panel.pairs_used")` consecutive-year pairs contributed by
`r mw_n("panel.providers_used")` providers. Over a career, spanning a median of
`r mw_safe_stat("panel.median_span")` years between first and last observation,
`r mw_safe_stat("persist.career_county", "%.1f")`% of providers ended in the
county where they began (Table 2).

The county estimate falls between the ZIP-code estimate below it and the state
estimate above it at both horizons, as it must, which is a consistency check on
all three. Career county persistence
(`r mw_safe_stat("persist.career_county", "%.1f")`%) substantially exceeded
career ZIP persistence (`r mw_safe_stat("persist.career_zip", "%.1f")`%),
indicating that a large share of address changes do not cross a county line.
Studies using ZIP change as a mobility proxy will overstate relocation by
roughly `r sprintf("%.0f", K$persist$career_county - K$persist$career_zip)`
percentage points over a career.

## Rurality

Annual persistence differed little across origin strata: a remote-origin
midwife changed county in about
`r mw_safe_stat("rural.annual_remote_move", "%.1f")`% of years against
`r mw_safe_stat("rural.annual_metro_move", "%.1f")`% for a metropolitan-origin
midwife. Over a career the gradient was larger and monotone (Table 3,
Figure 1). Career
same-county persistence was
`r mw_pct_ci(K$rural$table$pct[1], K$rural$table$lo[1], K$rural$table$hi[1])`
for metropolitan origin,
`r mw_pct_ci(K$rural$table$pct[2], K$rural$table$lo[2], K$rural$table$hi[2])`
for nonmetropolitan adjacent, and
`r mw_pct_ci(K$rural$table$pct[3], K$rural$table$lo[3], K$rural$table$hi[3])`
for nonmetropolitan remote origin (`r mw_pval(K$rural$trend_p)` for trend). The
difference between the extreme strata was
`r mw_safe_stat("rural.diff", "%.1f")` percentage points (95% CI
`r mw_safe_stat("rural.diff_lo", "%.1f")`--`r mw_safe_stat("rural.diff_hi", "%.1f")`).

## Where movers go

Among providers who changed county over their career, destination was strongly
asymmetric (Table 4, Figure 2). Of nonmetropolitan-adjacent-origin movers,
`r mw_safe_stat("movers.adj_to_metro_pct", "%.0f")`% ended in a metropolitan
county, as did `r mw_safe_stat("movers.remote_to_metro_pct", "%.0f")`% of
remote-origin movers; `r mw_safe_stat("movers.metro_stay_pct", "%.0f")`% of
metropolitan-origin movers remained metropolitan.

Net flows were nevertheless close to balanced:
`r mw_n("movers.out_of_nonmetro")` nonmetropolitan-origin movers ended
metropolitan while `r mw_n("movers.into_nonmetro")` metropolitan-origin movers
ended nonmetropolitan. There is no large absolute rural drain in this cohort.
However, the nonmetropolitan base is small --
`r formatC(sum(K$rural$table$n[2:3]), format = "d", big.mark = ",")` providers
against `r formatC(K$rural$table$n[1], format = "d", big.mark = ",")`
metropolitan -- so an equal absolute flow represents a far larger share of the
rural workforce, and it is the rate of departure, not its count, that a county
experiences.

## Sensitivity analyses

Resolving county through the geocoding cascade and point-in-polygon assignment
rather than the ZCTA crosswalk reproduced the result from an independent
construction: annual persistence `r mw_safe_stat("geo.sens_annual", "%.1f")`%,
career persistence `r mw_safe_stat("geo.sens_career", "%.1f")`%, and a monotone
rural gradient of `r mw_safe_stat("geo.sens_metro", "%.1f")`%,
`r mw_safe_stat("geo.sens_adj", "%.1f")`%, and
`r mw_safe_stat("geo.sens_remote", "%.1f")`% across the three strata,
`r mw_safe_stat("geo.sens_gradient", "%.1f")` percentage points wide.

A third approach, using the Census batch geocoder alone, produced a visibly
flatter gradient (`r mw_safe_stat("geo.census_only_metro", "%.1f")`%,
`r mw_safe_stat("geo.census_only_adj", "%.1f")`%,
`r mw_safe_stat("geo.census_only_remote", "%.1f")`%). The cause is measurable:
the Census geocoder rejects rural route and post-office-box addresses and lost
`r mw_safe_stat("geo.census_only_loss_pct", "%.1f")`% of remote-origin
providers, precisely the stratum whose persistence is lowest. Restoring those
providers through the ArcGIS tier restores the gradient. Two methods with
unrelated failure modes agree, and the third disagrees for a reason that is
understood.

# Discussion

In a national cohort of certified midwives followed across
`r mw_safe_stat("panel.snapshots")` annual provider snapshots, geographic
persistence was high year to year and substantially lower across a career:
`r mw_safe_stat("persist.annual_county", "%.1f")`% of consecutive-year pairs
were in the same county, but only
`r mw_safe_stat("persist.career_county", "%.1f")`% of providers ended their
observed career where they began. Persistence declined monotonically with the
rurality of the county of first practice, with remote-origin midwives about
`r mw_safe_stat("rural.diff", "%.0f")` percentage points less likely than
metropolitan-origin midwives to end where they started.

These two findings carry different implications for the same policy argument,
and the horizon is what separates them. Over one to three years, retention is
geographically sticky enough that the incidence of a retention effect falls
approximately where the affected midwives already practice, and a county-level
incidence model is defensible without a mobility correction. Over a decade it
is not: with roughly
`r sprintf("%.0f", 100 - K$persist$career_county)`% of providers changing county
over a median `r mw_safe_stat("panel.median_span")` years, and
`r mw_safe_stat("rural.career_remote_move", "%.0f")`% of remote-origin
providers doing so, national headcount and county distribution decouple. A
model claiming that retaining midwives sustains supply in a specific county
over a decade needs a mobility term, and that term now has an empirical value
rather than an assumption.

The rurality gradient sharpens the point. The counties with the least obstetric
infrastructure are the counties whose midwives are least likely to
remain[@ref42], so the same national retention effect delivers less local
access where local access is scarcest. Because rural-origin movers
overwhelmingly relocate to metropolitan counties, a retention intervention that
is geographically untargeted will tend to consolidate the workforce where it is
already dense.

This analysis also illustrates a methodological hazard worth reporting. An
earlier version of this work, stratifying by each provider's *current* county
rather than county of origin, found no rural mobility penalty. That
specification counts a midwife who left a remote county as metropolitan and
attributes her move to metropolitan mobility; correcting it did not merely
attenuate the effect but reversed its direction. Any workforce mobility
analysis stratified on a time-varying geographic attribute is exposed to this
error[@todo_provider_mobility].

## Strengths and limitations

The principal strength is that the cohort is an enumeration rather than a
sample, followed annually for `r mw_safe_stat("panel.snapshots")` years with a
complete snapshot series, and that every input is public, so the analysis is
reproducible end to end from archived code and checksummed inputs. The county
outcome was resolved by two constructions with unrelated failure modes that
agree.

The principal limitation is that linkage is selected on an outcome correlated
with the exposure. Only `r mw_safe_stat("linkage.active_pct", "%.1f")`% of
ACTIVE certificants and `r mw_safe_stat("linkage.dead_pct", "%.1f")`% of
DECEASED certificants resolve, so providers who exited the workforce are
under-represented, and persistence among those we observe is not persistence
among all certificants. The direction of this bias is toward overstating
persistence.

That limitation binds the persistence estimates more tightly than the
composition ones, and the difference is worth stating rather than leaving to
the reader. Persistence cannot be bounded here at all: it requires repeated
observation, and the certificants outside the cohort have none. The
metropolitan share can be, and is -- `r mw_safe_stat("bounds.lower_pct", "%.1f")`%
to `r mw_safe_stat("bounds.upper_pct", "%.1f")`% roster-wide with no assumption
about the missingness mechanism (Appendix 1). More usefully, the missingness is
not entirely opaque: the `r mw_n("bounds.outside_n")` certificants outside the
cohort who could nonetheless be placed were
`r mw_safe_stat("bounds.outside_pct", "%.1f")`% metropolitan, marginally *less*
metropolitan than the cohort rather than more. If the certificants who could
not be placed resemble those who could, the cohort slightly overstates how
metropolitan the workforce is -- a bias in the same direction as the
persistence bias, and of a magnitude the tipping-point calculation shows the
rurality conclusion survives.

Second, practice address in the provider registry is the address a provider
maintains for enumeration, which may lag an actual relocation or reflect an
administrative rather than clinical site; both would bias persistence upward.
Third, `r mw_n("linkage.ambiguous")` records were deliberately left unresolved,
and if indistinguishable names are not distributed at random with respect to
mobility, the cohort is not representative of the roster. Fourth, county
assignment from ZIP is arbitrary for the `r mw_n("geo.arbitrary_zctas")` ZCTAs
whose dominant county holds under 60% of their land, although the sensitivity
analysis argues this does not drive the result. Finally, the population is
restricted to midwives certified by a single certifying body; midwifery care
delivered by certified professional midwives and state-licensed direct-entry
midwives is not represented, so midwifery access is understated, and unevenly
so across states[@todo_midwife_outcomes].

## Conclusion

Retention of certified midwives is geographically sticky over short horizons
and materially less so over a career, with the weakest persistence among
midwives who began practice in the most rural counties. Retention interventions
intended to improve local obstetric access should state their time horizon
explicitly, and should be targeted geographically rather than applied
uniformly, because the communities most likely to lose a midwife are those
least able to replace one.

\newpage

# Tables

```{r tbl1}
# TABLE 1 -- cohort characteristics.
#
# SELECTED, not dumped. artifacts/table1_midwives.csv carries 20 characteristic
# blocks, several of which (Medicare practice group size, telehealth, training
# institution) belong to other papers. The blocks named here are the ones a
# reader of a MOBILITY paper needs in order to judge who moved: credential,
# sex, age, time since certification, and the rurality the persistence analysis
# stratifies on. Named explicitly so adding a block to the artifact cannot
# silently lengthen Table 1.
BLOCKS <- c("Certification",
            "Sex",
            "Calibrated Age (100% Cohort Coverage)",
            "Years Since AMCB Initial Certification",
            "Rurality (RUCC 2023)")

t1src <- readr::read_csv(file.path(root, "artifacts", "table1_midwives.csv"),
                         show_col_types = FALSE, progress = FALSE)
t1src$category <- trimws(t1src$category)

rows <- list()
for (b in BLOCKS) {
  blk <- t1src[t1src$category == b, ]
  if (!nrow(blk)) next
  rows[[length(rows) + 1]] <- data.frame(
    Characteristic = paste0("**", b, "**"), `n (%)` = "", check.names = FALSE)
  for (i in seq_len(nrow(blk))) {
    pct <- blk$percent[i]
    rows[[length(rows) + 1]] <- data.frame(
      Characteristic = paste0("\u00a0\u00a0", blk$characteristic[i]),
      `n (%)` = if (is.na(pct)) {
        sprintf("%s (not recorded)", formatC(blk$n[i], format = "d", big.mark = ","))
      } else {
        sprintf("%s (%.1f)", formatC(blk$n[i], format = "d", big.mark = ","), pct)
      },
      check.names = FALSE)
  }
}
t1out <- do.call(rbind, rows)

knitr::kable(t1out, row.names = FALSE, align = "lr",
  caption = sprintf("Table 1. Characteristics of the %s actively certified, primary-linked midwives. Values are n (%%) unless noted. Percentages within a block are computed on the members of that block with the characteristic recorded, so a block containing a not-recorded row does not sum to 100.",
                    formatC(t1src$n[1], format = "d", big.mark = ",")))
```

```{r tbl2}
t2 <- data.frame(
  Horizon = c("Year to year", "First vs last observation"),
  `Same county` = sprintf("%.1f", c(K$persist$annual_county, K$persist$career_county)),
  `Same ZIP code` = sprintf("%.1f", c(K$persist$annual_zip, K$persist$career_zip)),
  `Same state` = sprintf("%.1f", c(K$persist$annual_state, K$persist$career_state)),
  check.names = FALSE)
knitr::kable(t2, row.names = FALSE, align = "lrrr",
  caption = "Table 2. Persistence at two horizons, by geographic unit. Values are percentages.")
```

```{r tbl3}
r <- K$rural$table
t3 <- data.frame(
  `Origin stratum` = r$origin,
  Providers = formatC(r$n, format = "d", big.mark = ","),
  `Same county, %` = sprintf("%.1f", r$pct),
  `95% CI` = sprintf("%.1f-%.1f", r$lo, r$hi),
  check.names = FALSE)
knitr::kable(t3, row.names = FALSE, align = "lrrr",
  caption = sprintf("Table 3. Career same-county persistence by rurality of the county of first observed practice. %s for trend; metropolitan minus remote %.1f percentage points (95%% CI %.1f-%.1f).",
                    gsub("\\*", "", mw_pval(K$rural$trend_p)),
                    K$rural$diff, K$rural$diff_lo, K$rural$diff_hi))
```

```{r tbl4}
m <- K$movers$table
t4 <- data.frame(`Origin stratum` = rownames(m), check.names = FALSE)
for (cn in colnames(m)) t4[[cn]] <- formatC(m[, cn], format = "d", big.mark = ",")
knitr::kable(t4, row.names = FALSE, align = "lrrr",
  caption = sprintf("Table 4. Destination stratum of providers who changed county over their career. Net flow: %s nonmetropolitan-origin providers ended metropolitan; %s metropolitan-origin providers ended nonmetropolitan.",
                    formatC(K$movers$out_of_nonmetro, format = "d", big.mark = ","),
                    formatC(K$movers$into_nonmetro, format = "d", big.mark = ",")))
```


\newpage

# Figures

```{r fig1, fig.width=6.5, fig.height=3.4, fig.cap="Figure 1. Career same-county persistence by rurality of the county of first observed practice. Points are proportions; bars are 95% Wilson score intervals. The axis is restricted to the range of the data and does not begin at zero, so distances between points are comparable but the marks must not be read as areas."}
suppressPackageStartupMessages(library(ggplot2))

# FORM. Three ordered strata, one measure, uncertainty that matters -- the
# remote stratum has 520 providers and its interval is three times the width of
# the metropolitan one. That is a dot-and-whisker, not a bar chart: bars imply a
# zero baseline, and the informative range here is 55-72%, so a zero-based bar
# would compress the whole finding into the top eighth of the panel.
#
# GRAYSCALE BY CONSTRUCTION. Obstetrics & Gynecology prints many figures without
# color. There is one series, so no hue carries identity and nothing is lost.
r <- K$rural$table
r$origin <- factor(r$origin, levels = rev(r$origin))

ggplot(r, aes(x = pct, y = origin)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.10, linewidth = 0.5,
                 color = "grey35") +
  geom_point(size = 3.1, color = "grey15") +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -1.5, size = 3.5,
            color = "grey15") +
  scale_x_continuous(limits = c(55, 71), breaks = seq(55, 70, 5),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Ended career in county of first practice", y = NULL) +
  theme_classic(base_size = 11) +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey92", linewidth = 0.3),
        plot.margin = margin(6, 14, 6, 6))
```

```{r fig2, fig.width=6.5, fig.height=3.2, fig.cap="Figure 2. Destination of providers who changed county over their career, as a percentage of movers from each origin stratum. Counts are given in Table 4. Rural-origin movers relocate predominantly to metropolitan counties; metropolitan-origin movers overwhelmingly remain metropolitan."}
# FORM. The counts in Table 4 span 36 to 3,868, so a heatmap or a count bar
# would show one dark cell and eight pale ones -- the magnitude swamps the
# pattern. The FINDING is compositional: where movers from each origin end up.
# So the measure is the row percentage, and 100% stacked bars carry it.
#
# The fill is an ordered gray ramp because the destination strata are ordered
# (metropolitan -> adjacent -> remote), which is a sequential encoding, not a
# categorical one. Every segment wide enough to hold one is directly labeled,
# so identity never rests on the fill alone.
m <- K$movers$table
d <- data.frame(
  origin = rep(rownames(m), times = ncol(m)),
  dest   = rep(colnames(m), each = nrow(m)),
  n      = as.vector(m))
d$origin <- factor(d$origin, levels = rev(rownames(m)))
d$dest   <- factor(d$dest, levels = colnames(m))
d$pct    <- 100 * d$n / ave(d$n, d$origin, FUN = sum)

ggplot(d, aes(x = pct, y = origin, fill = dest)) +
  geom_col(width = 0.62, color = "white", linewidth = 0.7,
           position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = ifelse(pct >= 7, sprintf("%.0f%%", pct), ""), color = dest),
            position = position_stack(vjust = 0.5, reverse = TRUE), size = 3.3,
            show.legend = FALSE) +
  scale_fill_manual(values = c("To metropolitan"      = "grey30",
                               "To nonmetro adjacent" = "grey65",
                               "To nonmetro remote"   = "grey86"),
                    name = NULL) +
  scale_color_manual(values = c("To metropolitan"      = "white",
                                "To nonmetro adjacent" = "grey15",
                                "To nonmetro remote"   = "grey15")) +
  scale_x_continuous(expand = c(0, 0), labels = function(x) paste0(x, "%")) +
  labs(x = "Percentage of movers from that origin", y = NULL) +
  theme_classic(base_size = 11) +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(0.42, "cm"),
        plot.margin = margin(6, 14, 6, 6))
```

```{r fig3, fig.width=7.2, fig.height=4.6, fig.cap="Figure 3. Certified nurse-midwives and certified midwives per 1,000 births, by state. Alaska, Hawaii and the territories are omitted from the projection. Territories are also excluded from the color scale: Guam records two midwives against zero births, a rate of 2,000 per 1,000, which would compress all 51 mapped jurisdictions into a single class."}
# Built with {mysterymaps} (github.com/mufflyt/mysterymaps), the canonical map
# library for this group. R/lib/mysterymaps_dep.R records why nothing here
# re-implements a scale or a legend locally.
#
# SCOPE IS THE 50 STATES AND DC, which is the manuscript's population, not a
# cosmetic filter. The territories carry rates computed on zero births.
sw <- readr::read_csv(file.path(root, "artifacts", "state_obstetric_workforce.csv"),
                      show_col_types = FALSE, progress = FALSE) |>
  dplyr::filter(.data$state %in% c(state.abb, "DC"), .data$births > 0)

# print() is load-bearing: the builder returns invisible(p), so without it
# knitr runs the chunk, draws nothing, and leaves a caption over blank space.
print(mysterymaps::mysterymaps_geographic_map(
  data        = as.data.frame(sw),
  state_col   = "state",
  outcome_col = "midwives_per_1k_births",
  fill_label  = "Midwives per\n1,000 births",
  palette     = "viridis",
  low_states_warn = 0L,
  # The default legend formatter is scales::percent_format(), correct only for a
  # proportion on [0, 1]. It rendered this column's 0.53-10.79 as "250%" to
  # "1 000%". The `labels` argument was added to the library for this.
  labels      = function(x) sprintf("%.0f", x)))
```

```{r fig4, fig.width=7.2, fig.height=4.6, fig.cap="Figure 4. Midwife share of the identified birth-attendant workforce, by state. The denominator is midwives plus generalist obstetrician-gynecologists plus maternal-fetal medicine subspecialists; it excludes family physicians and certified professional midwives, both of which attend births, so the share is an upper bound on midwives' share of that larger workforce."}
print(mysterymaps::mysterymaps_geographic_map(
  data        = as.data.frame(dplyr::mutate(sw,
                  share_pct = 100 * .data$midwife_share_of_attendants)),
  state_col   = "state",
  outcome_col = "share_pct",
  fill_label  = "Midwife share of\nbirth attendants",
  palette     = "cividis",
  low_states_warn = 0L,
  labels      = function(x) paste0(sprintf("%.0f", x), "%")))
```

\newpage

# References

::: {#refs}
:::
